Covariate analysis of MSBB microarray data

### 2015-12-31 code/Rmd/MSSM_Array.Rmd
Major modifications in the MSSM microarray Rmd file

---
title: "Differential expression analysis of MSBB microarray data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn3721177';
activityName = 'Differential expression analysis';
activityDescription = 'Differential expression analysis based on image features';

thisFileName <- 'DiffExpAnal_MSSM_MA.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Takeda_contract", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('Rmd/',thisFileName))
```
### Image Analysis
1. Image features used:
  * FreeSurfer.thickness..median  
  * mean.curvature..median         
  * Volume
  * travel.depth..median  
  * FreeSurfer.convexity..sulc...median
```{r load.data}
adni.mem.1_5T.3Y <- read.csv('./adni.mem_ranked_1.5T_3Yr_takeda_4_6_15.csv')
label.name.conversion <- read.csv('./label_names_conversion.csv')

adni.mem.1_5T.3Y <- dplyr::mutate(adni.mem.1_5T.3Y, 
                                  sig.feature.all = a.ks.pvals <= 0.05 & a.cor.pvals <= 0.05,
                                  sig.feature.aks = a.ks.pvals <= 0.05,
                                  feature = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[-(length(x))],collapse='.')}),
                                  region = sapply(str_split(adni.mem.1_5T.3Y$var.nms,'\\.'),function(x){paste(x[(length(x))],collapse='.')}))
adni.mem.1_5T.3Y <- merge(adni.mem.1_5T.3Y,label.name.conversion, by.x = 'region', by.y = 'Label')
sig.table.all <- xtabs(sig.feature.all ~ feature + region,data = adni.mem.1_5T.3Y)
colnames(sig.table.all) <- label.name.conversion[match(label.name.conversion$Label,colnames(sig.table.all)),1]

unique.features <- unique(sapply(str_split(rownames(sig.table.all),'\\.'),function(x){x <- x[x!=""];l <- length(x);ifelse(l>1,paste(x[-(l)],collapse='.'),x)}))

region.ranking <- colSums(sig.table.all[rownames(sig.table.all) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]

unique.regions.correlated <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

sig.table.aks <- xtabs(sig.feature.aks ~ feature + Label.name, data = adni.mem.1_5T.3Y)
region.ranking <- colSums(sig.table.aks[rownames(sig.table.aks) %in% c('FreeSurfer.thickness..median','mean.curvature..median','Volume','travel.depth..median','FreeSurfer.convexity..sulc...median'),])
region.ranking <- sort(region.ranking,decreasing = T)
region.ranking <- region.ranking[region.ranking!=0]
unique.regions <- unique(sapply(str_split(names(region.ranking),'-'),function(x){x[[length(x)]]}))

```
2. Significantly changed regions are: `r unique.regions`
3. Significantly changed regions that are also correlated with ADNI.MEM are: `r unique.regions.correlated`
4. Following brain regions were analysed for differential expression between case and control

@@

```{r differential.expression}
# Function to calculate differnetial expression
differential.expression <- function(EXPR_ID, COV_ID){

  # Download expression data
  EXPR_OBJ <- synGet(EXP_ID)
  EXPR <- read.table(EXPR_OBJ@filePath,header=T,sep='\t',check.names = F)
  EXPR <- EXPR[EXPR$Gene.Symbol!='',]
  rownames(EXPR) <- EXPR$ID
  ANNOTATIONS <- dplyr::select(EXPR,ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID)
  EXPR <- dplyr::select(EXPR,-c(ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID))

  Collapsed.obj <- collapseRows(EXPR,ANNOTATIONS$Gene.Symbol,rownames(EXPR))
  EXPR <- Collapsed.obj$datETcollapsed

  # Download covariates data
  COV_OBJ <- synGet(COV_ID)
  COV <- fread(COV_OBJ@filePath,sep='\t',header=T,data.table=F)
  rownames(COV) <- COV$BrainBank
  COV <- dplyr::select(COV,-(BrainBank))
  COV <- dplyr::mutate(COV,Dx = ifelse(CDR<1,'Control','AD'))

@@

  # Match expression to covariates data
  ind <- match(colnames(EXPR),rownames(COV))
  EXPR <- EXPR[,!is.na(ind)]
  COV <- COV[ind[!is.na(ind)],]

  # Convert covariates to factors and numeric variables accordingly
  COV[,FactorCovariates] <- lapply(COV[,FactorCovariates], factor)  
  COV[,ContCovariates] <- lapply(COV[,ContCovariates], as.numeric)

  # Find PC of gene expression and significant covariates that are highly correlated with PCs
  preAdjustedSigCovars <- runPCAandPlotCorrelations(EXPR, COV,'expression', isKeyPlot=TRUE)

@@

  # Get design matrix
  DESIGN <- getDesignMatrix(COV[,'Dx',drop=F], Intercept = F)
  print(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))

#   # Calculate correlation between donors
#   correlation <- duplicateCorrelation(EXPR, design=DESIGN$design, block=ANNOTATIONS$Gene.Symbol)
#   print(paste('Consensus correlation is', correlation$consensus.correlation))
#   
  # Fit linear model
  FIT <- lmFit(EXPR, design = DESIGN$design)

  # Calculate residuals matrix
  RESIDUAL <- residuals.MArrayLM(FIT,EXPR)

  # Find correlation between residuals and covariates 
  postAdjustedSigCovars <- runPCAandPlotCorrelations(RESIDUAL, COV,'residual expression', isKeyPlot=TRUE)

@@

  # Make contrast to check differential expression
  CONT.NAMES <- expand.grid(colnames(DESIGN$design)[2],colnames(DESIGN$design)[1])
  CONT.NAMES <- apply(CONT.NAMES,1,paste,collapse ='-')

  CONT <- makeContrasts(contrasts=CONT.NAMES,levels=colnames(FIT$coefficients))

  # Refit contrasts
  FIT.CONTRAST <- contrasts.fit(FIT,CONT)

  # Estimate moderated t-statistics
  FIT.CONTRAST <- eBayes(FIT.CONTRAST)

  # Obtain all the differential expession combinations
  DEXP <- list()
  DEXP$logFC <- data.frame(row.names = rownames(EXPR))
  DEXP$adj.P.Val <- data.frame(row.names = rownames(EXPR))

  for (i in colnames(CONT)){
    tmp <- topTable(FIT.CONTRAST, coef=i, number=dim(EXPR)[1])    
    DEXP$logFC[,i] <- tmp[rownames(DEXP$logFC),'logFC']
    DEXP$adj.P.Val[,i] <- tmp[rownames(DEXP$adj.P.Val),'adj.P.Val']
  }

  # Get number of significantly differentialy expressed genes with adj.P.Val <= 0.05 and abs(logFC) >= 0
  DEXP$SIG.EXP <- DEXP$adj.P.Val<=0.05 #& abs(DEXP$logFC) >= 0   
  DEXP$NUM.SIG.EXP <- colSums(DEXP$SIG.EXP)

  return(list(FIT = FIT, FIT.CONTRAST = FIT.CONTRAST, DEXP = DEXP, ANNOTATIONS = ANNOTATIONS))
}
```
### Middle Temporal Analysis
```{r middle.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191101'
COV_ID <- 'syn3205399'  
ALL_USED_IDs <- unique(c(EXP_ID,COV_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
FC <- DIFF.EXP$DEXP$logFC
colnames(FC) <- paste('MSSM.MA.MiddleTemporal',colnames(FC),sep='.')
FC <- rownameToFirstColumn(FC,'hgnc_symbol')

PVAL <- DIFF.EXP$DEXP$adj.P.Val
colnames(PVAL) <- paste('MSSM.MA.MiddleTemporal',colnames(PVAL),sep='.')
PVAL <- rownameToFirstColumn(PVAL,'hgnc_symbol')
```
### Superior Temporal Analysis 
```{r superior.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191103'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.SuperiorTemporal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Inferior Temporal Analysis 
```{r inferior.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191099'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.InferiorTemporal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Precuneus Superior Parietal Lobule Analysis
```{r superior.parietal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191119'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.PrecuneusSuperiorParietal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Caudal Anterior Cingulate Analysis
```{r anterior.cingulate, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191107'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.CaudalAnteriorCingulate',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Hippocampal Analysis
Hippocampal region shows shape changes but not correltated with ADNIMEM
```{r hippocampal, cache=TRUE, fig.height=10, fig.width=15}
# Print significant features
print('Significantly changed image features of hippocampal region are:')
tmp <- filter(adni.mem.1_5T.3Y, a.ks.pvals <= 0.05 & Label.name %in% grep('hippocampal',adni.mem.1_5T.3Y$Label.name,value=T))
kable(tmp)

# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191109'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.Hippocampal',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Frontal Pole Analysis
Frontal pole analysis was performed (since this is one of the regions for which RNASeq is available from MSSM)
```{r frontal.pole, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- 'syn3191107'
ALL_USED_IDs <- unique(c(ALL_USED_IDs,EXP_ID))

# Obtain differential expession
DIFF.EXP <- differential.expression(EXPR_ID,COV_ID)

# Print results
tmp <- as.data.frame(DIFF.EXP$DEXP$NUM.SIG.EXP)
colnames(tmp) <- 'Diff exp probes at FDR less than 0.05'
kable(tmp)

# Get fold change and pvalues
tmp <- DIFF.EXP$DEXP$logFC
colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
FC <- join(FC, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')

tmp <- DIFF.EXP$DEXP$adj.P.Val
colnames(tmp) <- paste('MSSM.MA.FrontalPole',colnames(tmp),sep='.')
PVAL <- join(PVAL, rownameToFirstColumn(tmp,'hgnc_symbol'), by = 'hgnc_symbol', match = 'all')
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- File(paste0('./',thisFileName), name = "Differential Expression Analysis MSSM MA", parentId = parentId)
CODE <- synStore(CODE, used = ALL_USED_IDs, activityName = activityName, executed = thisFile, activityDescription = activityDescription)

# Fold change
FC <- FC[, unique(colnames(FC))]
write.table(FC, file='./FoldChange_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
FC_OBJ <- File('./FoldChange_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA logFC',parentId = parentId)
FC_OBJ <- synStore(FC_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)

# Pvalue
PVAL <- PVAL[, unique(colnames(PVAL))]
write.table(PVAL, file='./AdjustedPval_MSSM_MA.tsv', sep='\t', row.names=F, quote=F)
PVAL_OBJ <- File('./AdjustedPval_MSSM_MA.tsv',name = 'Differential Expression Analysis MSSM MA adjPval',parentId = parentId)
PVAL_OBJ <- synStore(PVAL_OBJ, used = ALL_USED_IDs, activityName = activityName, executed=CODE$properties$id, activityDescription = activityDescription)
```
|  *Results*           |  *SynapseID*                |
|  -------             |   ---------                 |
|  Log fold change     |  `r FC_OBJ$properties$id`       |
|  Adjusted pvalues    |  `r PVAL_OBJ$properties$id`     |

>>>>

---
title: "Covariate analysis of MSBB microarray data"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knit2synapse)
library(synapseClient)

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570248';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSBB microarrays';

thisFileName <- 'MSSM_Array.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
Following brain regions were analysed for differential expression between case and control

@@

```{r covariate analysis}
# Function to do covariates adjustments
covariatesAdjustments <- function(exprId, covId){
  # Download expression data
  expr <- read.table(synGet(exprId)@filePath, header=T, sep='\t', check.names = F)
  expr <- expr[expr$Gene.Symbol!='',]
  rownames(expr) <- expr$ID
  ANNOTATIONS <- dplyr::select(expr,ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID)
  expr <- dplyr::select(expr,-c(ID,GB_ACC,Gene.Symbol,ENTREZ_GENE_ID))
  rownames(expr) = ANNOTATIONS$ID

  Collapsed.obj <- collapseRows(expr,ANNOTATIONS$Gene.Symbol,ANNOTATIONS$ID)
  expr <- Collapsed.obj$datETcollapsed

  # Download covariates data
  covar <- fread(synGet(covId)@filePath,sep='\t',header=T,data.table=F)
  BrainBank <- covar$BrainBank
  covar <- dplyr::select(covar,-(BrainBank))
  covar <- dplyr::mutate(covar,Dx = ifelse(CDR<1,'Control','AD'))
  rownames(covar) <- BrainBank

@@

  # Match expression to covariates data
  ind <- match(colnames(expr),rownames(covar))
  expr <- expr[,!is.na(ind)]
  covar <- covar[ind[!is.na(ind)],]

  # Convert covariates to factors and numeric variables accordingly
  covar$Age = gsub('\\+','',covar$Age)
  covar[,FactorCovariates] <- lapply(covar[,FactorCovariates], as.factor)
  covar[,ContCovariates] <- lapply(covar[,ContCovariates], as.numeric)

  # Find PC of gene expression and significant covariates that are highly correlated with PCs
  covar.tmp = covar
  covar.tmp[,FactorCovariates] <- lapply(covar.tmp[,FactorCovariates], as.numeric)
  preAdjustedSigCovars <- runPCAandPlotCorrelations(expr, covar.tmp, 'expression', isKeyPlot=TRUE)

@@

  # Get design matrix
  DESIGN <- getDesignMatrix(covar[,'Dx',drop=F], Intercept = F)
  print(paste('Fitting linear model with the following coefficients:',paste(colnames(DESIGN$design),collapse=',')))

  # Fit linear model
  fit <- lmFit(expr, design = DESIGN$design)

  # Calculate residuals matrix
  residualFit <- residuals.MArrayLM(fit, expr)
  residualFit = residualFit + fit$coefficients[,c('DxAD','DxControl')] %*% t(DESIGN$design[,c('DxAD','DxControl')])

  # Find correlation between residuals and covariates 
  postAdjustedSigCovars <- runPCAandPlotCorrelations(residualFit, covar.tmp,'residual expression', isKeyPlot=TRUE)

@@

  return(list(EXPR = expr,
              COVAR = covar,
              FIT = fit, 
              RESIDUAL = residualFit, 
              DESIGN = DESIGN, 
              ANNOTATIONS = ANNOTATIONS))
}
```
### Middle Temporal Analysis
```{r middle.temporal, cache=TRUE, fig.height=10, fig.width=15}
# SynIDs of expression and covariates matrix
EXP_ID <- c('Middle Temporal' = 'syn3191101',
            'Superior Temporal' = 'syn3191103',
            'Inferior Temporal' = 'syn3191099',
            'Superior Parietal Lobule' = 'syn3191119',
            'Caudal Anterior Cingulate' = 'syn3191107',
            'Parahippocampal Gyrus' = 'syn3191109',
            'Frontal Pole' = 'syn3191107')
COV_ID <- 'syn3205399'

# Obtain adjusted expession
adjustedExpression <- lapply(EXP_ID, covariatesAdjustments, COV_ID)
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE}
# Code
CODE <- Folder(name = "MSBB Microarray", parentId ="syn5570248")
CODE <- synStore(CODE)

ALL.OBJ = mapply(function(exp, regionName, parentId, used, executed, activityName, activityDescription){
  # Expression
  expr = rownameToFirstColumn(exp$EXPR, 'hgnc_symbols')
  write.table(expr, file='expression.tsv', sep = '\t', row.names=F, quote=F)
  obj = File('expression.tsv', 
             name = paste('Expression',regionName), 
             parentId = parentId)
  obj = synStore(obj,
             used = used, 
             activityName = activityName, 
             executed = executed, 
             activityDescription = activityDescription)

  # Covariates
  expr = rownameToFirstColumn(exp$COVAR, 'SampleId')
  write.table(expr, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
  obj = File('covariates.tsv', 
             name = paste('Covariates',regionName), 
             parentId = parentId)
  obj = synStore(obj,
             used = used, 
             activityName = activityName, 
             executed = executed, 
             activityDescription = activityDescription)
},adjustedExpression, names(adjustedExpression), 
MoreArgs = list(CODE$properties$id, ALL_USED_IDs, thisFile, activityName, activityDescription), 
SIMPLIFY = F)       
```
